---
title: "Fitting a logistic regression model with Rstan"
author: "Georgia Tsambos"
date: "09/01/2019"
header-includes: \usepackage{bbm}
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

I'm going to attempt to fit a logistic regression model to one of the ACTN3 studies *(Yang 2003)* using RStan.

The dataset is as follows:

|              | **RR**  | **RX**  | **XX** | *Total* |
|--------------|---------|---------|--------|---------|
| **Athletes** | 53      | 48      | 6      | *107*   |
| **Controls** | 130     | 226     | 80     | *436*   |
| *Total*      | *183*   | *274*   | *86*   | *543*   |

## The Rstan model

To design the rstan model for this dataset, let's consider what we know about it and the analysis we wish to perform.

Firstly, we have an equation for the regression to perform:

\[
\mathrm{logit}(p_{i}) = 
\log \left( \dfrac{P(\mathrm{athlete}\mid G_{i})}{P(\mathrm{control}\mid G_{i})} \right)\tag{1} = 
\mu + \beta G_{i} + \gamma \cdot I(G_{i} = 1)
\]

The parameters of interest that we wish to estimate are the regression coefficients $\mu, \beta, \gamma$.

Thus we have RStan chunk
```
parameters {
  real mu;        // baseline odds of being an athlete
  real beta;      // additive effect
  real gamma;     // dominance effect
}
```
Each of the $n=543$ individuals in the study is associated with a known predictor variable $g_i \in \{ XX=0,\thinspace RX=1,\thinspace RR=2\}$ and a known case/control status $y_i \in \{ \mathrm{athlete}=1, \mathrm{control=0} \}$ for $i=1,\ldots,n$. There is also an indicator $I(G_1 = 1)$.

These all enter the RStan code as `data`:

Note that I define `g` and `ind` as vectors to make it possible to use certain operations on them later on.
```
data {
  int<lower=0> N;               // number of individuals
  vector[N] g;    // genotype
  int<lower=0,upper=1> y[N];    // case/control status
  vector[N] ind;  // indicator for heterozygotes
}
```
The case-control status $y_i$ of individual $i$ has distribution $\mathrm{Bernoulli}(p_i)$, where
\[
p_i = \mathrm{logit}^{-1} \left( \mu + \beta G_i + \gamma\cdot I(G_i=1)  \right).\tag{2}
\]
This relationship is entered into the `model` chunk:
```
model {
  y ~ bernoulli_logit(mu + beta * g + gamma * ind);
}
```
Next, we consider the priors on the variables of interest $\beta, \gamma$.
We'll specify a very general multivariate normal prior based on a normal distribution (this will be more meaningful when we extend to a meta-analysis involving more than one study!)
\[
\begin{bmatrix}
\beta \\ \gamma
\end{bmatrix} \sim
N \left(
\begin{bmatrix}
\beta_0, \gamma_0
\end{bmatrix}, \thinspace
\begin{bmatrix}
\tau_\beta^2 & \rho \tau_\beta \tau_\gamma \\
\rho \tau_\beta \tau_\gamma & \tau_\gamma^2
\end{bmatrix}
\right)\tag{5}
\]

```
model {
  y ~ bernoulli_logit(mu + beta * g + gamma * ind);
  {
    vector[2] coeffs = [ beta, gamma ]';
    vector[2] coeffs0 = [ beta0, gamma0 ]';
    matrix[2,2] sigma = [ [ taubeta*taubeta, rho*taubeta*taugamma ], [ rho*taubeta*taugamma,  taugamma*taugamma ] ];
    coeffs ~ multi_normal(coeffs0, sigma);
  }
}
```

Note that this introduces new hyperparameters $\beta_0, \gamma_0, \tau_\beta, \tau_\gamma, \rho$ (these will be more meaningful later with the full meta-analysis!) These should be added to the `parameter` chunk:
```
parameters {
  real mu;        // baseline odds of being an athlete
  real beta;      // additive effect
  real gamma;     // dominance effect
  // hyperparameters:
  real beta0;
  real gamma0;
  real<lower=0> taubeta;
  real<lower=0> taugamma;
  real<lower=-1,upper=1> rho;
}
```
As we are assuming the default weakly informative uniform prior on $\mu$, this distribution need not be specified in the script.

### The full RStan model

I save this in a script called `01.single-logit-model.stan`.

```
data {
  int<lower=0> N;               // number of individuals
  vector[N] g;    // genotype
  int<lower=0,upper=1> y[N];    // case/control status
  vector[N] ind;  // indicator for heterozygotes
}
parameters {
  real mu;        // baseline odds of being an athlete
  real beta;      // additive effect
  real gamma;     // dominance effect
  // hyperparameters:
  real beta0;
  real gamma0;
  real<lower=0> taubeta;
  real<lower=0> taugamma;
  real<lower=-1,upper=1> rho;
}
model {
  y ~ bernoulli_logit(mu + beta * g + gamma * ind);
  {
    vector[2] coeffs = [ beta, gamma ]';
    vector[2] coeffs0 = [ beta0, gamma0 ]';
    matrix[2,2] sigma = [ [ taubeta*taubeta, rho*taubeta*taugamma ], [ rho*taubeta*taugamma,  taugamma*taugamma ] ];
    coeffs ~ multi_normal(coeffs0, sigma);
  }
}

```

## Fitting the model in R

First, we need to load RStan. I also enable parallel processing since I'm working on my own computer (for now).

```{r loadpackages, echo = T, results = 'hide'}
library(rstan, quietly = T)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

Next, let's enter the data. 
We could have set up the model above in a way that allows us to enter the totals, rather than just each individual separately.
Let's maybe explore that later.

I'll enter the data in the columns going across rows left to right, then down columns:

```{r defineData, echo = T, results = 'hide'}
# Number of study individuals
N <- 543

# Case control status
y <- c(rep(1, 107), rep(0, 436))

# Genotype
g <- c(rep(0, 53), rep(1, 48), rep(2, 6), rep(0, 130), rep(1, 226), rep(2, 80))

# Indicator status
ind <- sapply(g, function(x) abs(x - 1))
```

Let's first fit a regular glm to this

```{r glmFit}
fit0_data <- data.frame(phenotype = y, additive = g, dominance = ind)
fit0 <- glm(phenotype ~ additive + dominance, data = fit0_data, family = 'binomial')
summary(fit0)
```
As expected from a quick visual 'look' at the data, there is a negative association between the X allele and the probability of being an athlete. The dominance coefficient is negative (suggesting recessiveness), though insignificant.

Next, let's fit the Stan model using the following (simplified) model:
```
data {
  int<lower=0> N;               // number of individuals
  vector[N] g;    // genotype
  int<lower=0,upper=1> y[N];    // case/control status
  vector[N] ind;  // indicator for heterozygotes
}
parameters {
  real mu;        // baseline odds of being an athlete
  real beta;      // additive effect
  real gamma;     // dominance effect
}
transformed parameters {
  vector[N] p = mu + beta * g + gamma * ind;
}
model {
  y ~ bernoulli_logit(p);
}

```

```{r modelfit}
study_data <- list(N=N, g=g, y=y, ind=ind)
fit <- stan(file = '01.single-logit-model.stan', data = study_data)
```

Let's examine this object.
```{r examineFit}
print(fit)
```

First observation from this is: defining `p` as a `parameter` is a bad idea, because then the model attempts to fit each of these separately and there's a degrees of freedom problem. We can get around this with the followin?
```
data {
  int<lower=0> N;               // number of individuals
  vector[N] g;    // genotype
  int<lower=0,upper=1> y[N];    // case/control status
  vector[N] ind;  // indicator for heterozygotes
}
parameters {
  real mu;        // baseline odds of being an athlete
  real beta;      // additive effect
  real gamma;     // dominance effect
}
model {
  y ~ bernoulli_logit(mu + beta * g + gamma * ind);
}

```

Let's try again.


```{r modelfit1}
study_data <- list(N=N, g=g, y=y, ind=ind)
fit <- stan(file = '01.single-logit-model.stan', data = study_data)
```

Let's examine this object.
```{r examineFit1}
print(fit)
```
Note that these estimates are pretty similar to those obtained by `glm()`. Yay!

Next, let's try to fit the slightly more complicated model where there are hyperparameters governing the distributions of $\beta, \gamma$.

```
data {
  int<lower=0> N;               // number of individuals
  vector[N] g;    // genotype
  int<lower=0,upper=1> y[N];    // case/control status
  vector[N] ind;  // indicator for heterozygotes
}
parameters {
  real mu;        // baseline odds of being an athlete
  real beta;      // additive effect
  real gamma;     // dominance effect
  // hyperparameters:
  real beta0;
  real gamma0;
  real<lower=0> taubeta;
  real<lower=0> taugamma;
  real<lower=-1,upper=1> rho;
}
model {
  y ~ bernoulli_logit(mu + beta * g + gamma * ind);
  vector[2] coeffs = [ beta, gamma ]';
  vector[2] coeffs0 = [ beta0, gamma0 ]';
  matrix[2,2] sigma = [ [ taubeta*taubeta, rho*taubeta*taugamma ], [ rho*taubeta*taugamma, taugamma*taugamma ] ]
  coeffs ~ multi_normal(coeffs0, sigma)
}

```

```{r modelfit2}
fit <- stan(file = '01.single-logit-model.stan', data = study_data)
```
Let's examine this object.
```{r examineFit2}
print(fit)
```
Note that the estimated values of $\beta,\gamma$ are pretty decent, but $\beta_0,\gamma_0$ are way out of whack. I think this is possibly due to the fact that we're only observing a single value drawn from this distribution, and the fact that the prior was deliberately vague. Let's see if this improves with more studies.